%% This is the abstract of my thesis.
{\sis

Transactions are gaining ground as a programmer-friendly means of
expressing concurrency, as microarchitecture trends make it
clear that parallel systems are in our future.  This dissertation
presents the design and implementation of four efficient and powerful
transaction systems: \apex, an object-oriented software-only system;
UTM and LTM, two scalable systems using custom processor extensions;
and \hyx, a hybrid of the software and hardware systems,
obtaining the benefits of both.

The software transaction system implements \textit{strong atomicity},
which ensures that transactions are protected from the influence of
nontransactional code.  Previous software systems use weaker
atomicity guarantees because strong atomicity is presumed to be too
expensive.  In this thesis strong atomicity is obtained with
\punt{as little as 15\%}
minimal slowdown for nontransactional code.  Compiler analyses can
further improve the efficiency of the mechanism, which has been
formally verified with the \Spin model-checker.

The low overhead of \apex allows it to be profitably
combined with a hardware transaction system to provide fast execution
of short and small transactions, while allowing fallback to software
for large or complicated transactions.  I present UTM, a hardware
transactional memory system allowing unbounded virtualizable
transactions, and show how a hybrid system can be obtained.

}

% LocalWords:  analyses microarchitecture virtualizable UTM
