% -*- latex -*- This is a LaTeX document.
% $Id: quads.tex,v 1.18 1999-07-14 11:28:59 cananian Exp $
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[11pt,notitlepage,twocolumn,twoside]{article}
\usepackage{pdffonts} % PDF-friendly fonts.
\usepackage{comdef}

% What's the name of this project this week?
\newcommand{\Harpoon}{Harpoon}%{ActiveWorld}
\newcommand{\harpoon}{harpoon}%{activeWorld}

\author{C.~Scott~Ananian}
\title{\Harpoon{} Project Compiler Intermediate Representation}
\date{\today \\ $ $Revision: 1.18 $ $}

\begin{document}
\bibliographystyle{alpha}
\maketitle

\section{Conceptual Overview}

The \Harpoon{} project compiler front-end translates Java bytecode files
into a class-oriented intermediate representation which is intended to
be easier to analyze and manipulate than bytecode assembly language.
The intermediate representation is control-flow-graph structured, with
all control flow explicit.  It is also maximally factored and in
static single assignment (SSA) form.  Internally the intermediate
representation is known as ``QuadSSA,'' referring
to its derivation from quadruple-style IRs and its SSA form.

\subsection{Quadruples}
Unlike expression-tree structured intermediate representations, where
every operand can optionally be the root of an expression, quadruple
representations are flat.  The ``typical'' statement is of the form
$a \leftarrow b \oplus c$; the name \textit{quadruple} comes from the fact
that there are four components ($a$, $b$, $c$, $\oplus$)
\cite{appel:modern}.\footnote{We use $\oplus$ to stand for an arbitrary
binary operator.}  Obviously an IR expressive enough to represent the
entire Java language needs more than a four component operation
statement, but we have attempted to retain the atomic simplicity
of the quadruple form.

\subsection{Static Single-Assignment Form}

Quoting from Appel in \cite{appel:modern}:
\begin{quote}
Many dataflow analyses need to find the use-sites of each defined
variable, or the definition-sites of each variable used in an
expression.  The \textit{def-use chain} is a data structure that makes
this efficient: for each statement in the flow graph, the compiler can
keep a list of pointers to all the \textit{use} sites of variables
defined there, and a list of pointers to all \textit{definition} sites
of the variables used there\ldots .

An improvement on the idea of def-use chains is \textit{static
single-assignment form}, or \textit{SSA form}, an intermediate
representation in which each variable has only one definition in the
program text.  The one (static) definition-site may be in a loop that
is executed many (dynamic) times, thus the name \textit{static}
single-assignment form.
\end{quote}

An example of the use of SSA form is shown in figure \ref{fig:ssa1}.
Simple variable renaming suffices to transform straight-line code into
SSA form.  Subscripts are used to emphasize the relationship of the
renamed variables to the original variables.  A benefit of SSA form
which is obvious from the example is that unrelated uses of the same variable
in the source program ($\tt i_1$, $\tt i_2$) become different
variables in the SSA form, eliminating false dependencies.

\begin{figure}[t]
\begin{tabular}{lll}
Conventional    && Static Single Assignment \\ \hline
$\tt \ldots$    && $\tt \ldots$ \\
$\tt i=0$       && $\tt i_0 = 0$ \\
$\tt i=i+1$     && $\tt i_1 = i_0 + 1$ \\
$\tt j=func(i)$ && $\tt j_0 = func(i_1)$ \\
$\tt i=2$       && $\tt i_2 = 2$ \\
$\tt \ldots$    && $\tt \ldots$ \\
\\
\end{tabular}
\caption{SSA transformation of straight-line code.}
\label{fig:ssa1}
\end{figure}

SSA form becomes more complex when we
introduce branches and loops.  Figure \ref{fig:ssa2} shows the
necessary transformation.  You will notice the introduction of
\textit{phi functions} at locations where control flow merges.  The
\phifunction{} ``magically'' chooses a value from among its arguments
based on the control flow path used to reach it.  Note that, although
\phifunction{s} are impossible to implement directly in an instruction set
(due to their magical properties), they can be replaced by \texttt{move}
instructions along each control flow edge leading to the
\phifunction{}.  Doing so violates the static single assignment
constraints, but leads to code executable by real processors.

Unless you are implementing code generator backends, it is unlikely
you will need to so replace \phifunction{s} or view them as anything
but magical $n$-ary operators.  However, it is important to observe and 
maintain the ordering relationship between control-flow edges and 
\phifunction{} arguments during transformation and analysis.

\begin{figure}[t]
\begin{tabular}{lll}
Conventional    && Static Single Assignment \\ \hline
$\tt \ldots$    && $\tt \ldots$ \\
$\tt i=0$       && $\tt i_0 = 0$ \\
$\tt if$ $x$ $\tt then$ && $\tt if$ $x$ $\tt then$ \\
~~$\tt i=1$     && ~~$\tt i_1 = 1$ \\
                && $\tt i_2 = \phi(i_0, i_1)$ \\
$\tt j=func(i)$ && $\tt j_0 = func(i_2)$ \\
$\tt \ldots$    && $\tt \ldots$ \\
\end{tabular}
\caption{SSA transformation of branching code.}
\label{fig:ssa2}
\end{figure}

Analysis, transformation, and optimization of the
IR is simplified by its SSA form.
In addition, the QuadSSA form is \textit{maximally factored}.
Constants are not allowed as quadruple operands (except for a special
\texttt{const} operation); which creates a unique mapping from
variable names to values in the computation.  This simplifies value
analysis.

Appel describes several other benefits of the SSA form in \cite{appel:modern}:

\begin{quote}
%\begin{enumerate}
---%\item 
If a variable has $N$ uses and $M$ definitions (which occupy
about $N+M$ instructions in a program), it takes space (and time)
proportional to $N\cdot M$ to represent def-use chains---a quadratic
blowup.  For almost all realistic programs, the size of the SSA form
is linear in the size of the original program.

---%\item 
Uses and defs of variables in SSA form relate in a useful way to
the dominator structure of the control flow graph, which simplifies
algorithms such as interference-graph construction.
%\end{enumerate}
\end{quote}

\subsection{Exception handling}
Exception handling in the Java language complicates control-flow.
Operations (such as division, possibly by zero) may implicitly throw
exceptions that radically redirect the flow of control.  To facilitate
analysis, exception handling and its associated control-flow is made
explicit in the intermediate representation.  For example, null
pointer and array bounds checks are inserted before object and array
references, and division by zero is explicitly checked before every
division operation.  These explicit and comprehensive checks are
intended to be targeted by aggressive optimizations designed to
removed those cases with are redundant, impossible, or otherwise
unnecessary.  The design goal of the IR is that \textit{no}
statement should throw an implicit exception under any 
circumstance.
Pursuing that goal involves changing the semantics of method
invocation slightly: so that the \texttt{call} statement not throw an
implicit exception, the IR \texttt{call} has been defined to return
\textit{two} values.  In addition to the conventional (and optional)
method return value, an ``exception return value'' is defined to hold
the exception thrown by a method, or \texttt{null} if the method
completed without throwing an exception.  Explicit tests of the
exception return value can then be added after the \texttt{call}
statement, and control-flow made explicit as with the other IR
operations.  The \texttt{throw} statement in the IR is thus stripped
of its special meaning and becomes simply an alternate \texttt{return}
statement for the exception return value.\footnote{Try, catch, and
finally blocks are, of course, taken into consideration when a
bytecode \texttt{athrow} is translated to an IR \texttt{THROW}.}

\section{Implementation Details}

The IR described in these notes is defined in the Java package
\texttt{harpoon.IR.QuadSSA}.  Source code and binaries are available
at {\tt\small http://\bp www.magic.lcs.mit.edu/\bp Harpoon}.

The
QuadSSA statements are called \texttt{Quad}s and are subclasses of
\texttt{harpoon.IR.QuadSSA.Quad}; they are graph-structured and
doubly-linked to enable both forward and reverse traversal of the
control-flow graph.  Edges are represented by Objects to facilitate
associating analysis data with these control-flow graph edges.
The parent class \texttt{Quad} contains the
graph-oriented methods of the objects.  Its superinterface
\texttt{harpoon.ClassFile.HCodeElement} defines standard methods to
get object ID numbers and source file information which
are valid for elements of any intermediate representation.

An enumeration of
\texttt{Quad} types and their uses is provided in figure
\ref{fig:classes}.  It may be observed that the representation uses both
header and footer nodes, in the \texttt{HEADER} and \texttt{FOOTER}
classes.  \texttt{HEADER} nodes contain a special link to the
\texttt{FOOTER} to allow this node to be easily identified, and a
special subclass of \texttt{HEADER}, \texttt{METHODHEADER}, provides
information on the assignment of method arguments to compiler
temporary variables at the start of method code.

With the exception of \texttt{CJMP}, \texttt{SWITCH}, \texttt{PHI},
\texttt{HEADER}, \texttt{METHODHEADER}, and \texttt{FOOTER}, all
\texttt{Quad}s have exactly one predecessor and one successor in the
control flow graph.

\subsection{Quads}
Here are more details on each Quad statement.  First the header and
footer nodes:
\begin{description}
%
\item[HEADER($f$)] Start node in the control flow graph with end node $f$.
Performs no operation.  Zero predecessors, one successor.
%
\item[METHODHEADER($f$, $p$)] Start node in the control flow graph for
a method with parameters $p$ and end node $f$.  A subclass of \texttt{HEADER}.
The method arguments are loaded into $p_0 \ldots p_n$ before execution starts.
Zero predecessors, one successor.
%
\item[FOOTER( )] Final node in control-flow graph.  Performs no operation.
All \texttt{RETURN} and \texttt{THROW} statements must have the graph's
\texttt{FOOTER} as their only successor.  \texttt{FOOTER} nodes may
have any positive number of predecessors.  They have no successors.
\end{description}
%
The following \texttt{Quad}s modify the control-flow:
\begin{description}
%
\item[PHI($t$, $l$)] Control flow merges at \texttt{PHI} nodes.
A phi node represents a list of \phifunction{s} of the form
\[ t_i = \phi(l_{i0}, l_{i1} \ldots l_{ij}) \]
where $j$ is the \textit{arity} of the \phifunction; that is, 
the number of predecessors to the node.  Any non-negative number of
predecessors, one successor.
%
\item[CJMP($t$)] Conditional jump based on the boolean $t$.  If $t$ is
false (0), control flows to the first successor (\texttt{next[0]}).
If $t$ is true (1), control flows to the second successor
(\texttt{next[1]}).  One predecessor, two successors.
%
\item[SWITCH($t$, $k$)] Indexed jump.  Depending on the value of index
variable $t$ and a key list $k$, control is transferred to 
${target}_n$ where $t = k_n$.  If $t$ does not match any key in $k$,
control is transferred to the default ${target}_{n+1}$ where $k_n$ is
the last key in the key list.  One predecessor, multiple successors.
%
\item[RETURN($t$)] Return an optional value $t$ from this method.
One predecessor, one successor.  The single successor should be a
\texttt{FOOTER} node.
%
\item[THROW($t$)] Throws an exception $t$ as the result of this method.
One predecessor, one successor.  The single successor should be a
\texttt{FOOTER} node.
\end{description}
%
The remaining \texttt{Quad}s have no effect on control flow, and have
exactly one predecessor and one successor.  No exceptions are thrown.
\begin{description}
\item[AGET($t_1$, $t_2$, $t_3$)] Fetches the element at index $t_3$
from array $t_2$ and stores the value in $t_1$.
%
\item[ALENGTH($t_1$, $t_2$)] Puts the length of array $t_2$ into
variable $t_1$.
%
\item[ANEW($t_1$, $c$, $l$)] Creates a new uninitialized array with
type $c$ and dimensions $l_0 \ldots l_n$, storing a reference in
$t_1$.  The number of dimensions
supplied in list $l$ may be smaller than the number of dimensions of
array class type $c$, in which case only the $n$ dimensions specified
in $l$ will be created.
%
\item[ASET($t_1$, $t_2$, $t_3$)] Sets the element at index $t_2$ of
array $t_1$ to the value in $t_3$.
%
\item[CALL($m$, $t$, $p$, $r$, $e$)] Calls method $m$ of optional class
reference $t$ with parameter list $p_0 \ldots p_n$, 
putting the return value in $r$
if no exception is thrown, or setting $e$ to the thrown exception.
Either $r$ or $e$ will be \texttt{null} on completion of the
\texttt{CALL}.  Exception $e$ is not automatically thrown from the
method containing the \texttt{CALL}: $e$ must be explicitly tested and
its exception rethrown if that behavior is desired. $t$ is not needed
for static methods.
%
\item[COMPONENTOF($t_1$, $t_2$, $t_3$)] Puts the boolean value
\texttt{true} (1) in $t_1$ if object $t_3$ is an instance of the component
type of array $t_2$, or \texttt{false} (0) otherwise.
%
\item[CONST($t$, $c$, $y$)] Assigns numeric or string constant $c$ of 
type $y$ to variable $t$.
%
\item[GET($t_1$, $f$, $t_2$)] Puts the value of field $f$ of optional
object $t_2$ in variable $t_1$. $t_2$ is not necessary for static fields.
%
\item[INSTANCEOF($t_1$, $t_2$, $c$)] Puts the boolean value
\texttt{true} (1) in $t_1$ if object $t_2$ is an instance of class
$c$, or \texttt{false} (0) otherwise.
%
\item[MOVE($t_1$, $t_2$)] Copies the value in $t_2$ into $t_1$.
%
\item[NEW($t$, $c$)] Create a new uninitialized instance of class $c$,
storing a reference in $t$.  A class constructor must be invoked using
\texttt{CALL} in order to initialize the instance.
%
\item[NOP( )] Performs no operation.
%
\item[OPER($o$, $t$, $l$)] Performs operation $o$ on the variables in
list $l$, storing the result in $t$.  The operation is represented as
a string; figure \ref{fig:classes} lists all valid operation strings.
The operations performed by the strings are identical to the
operations performed by the Java bytecode operation of the same name,
except that no exceptions are ever thrown.  See \cite{jvm} for
details.
%
\item[SET($f$, $t_1$, $t_2$)] Sets field $f$ of optional object $t_1$ to the
value of $t_2$.  $t_1$ is not necessary for static fields.
\end{description}
%
The \texttt{harpoon.IR.QuadSSA.Code}
class provides a means to access the QuadSSA form of a given method;
see the definition of superclass \texttt{harpoon.ClassFile.HCode} and
the example code in \texttt{harpoon.Main.Main} for details.  

\newcommand{\cls}[1]{{\normalfont\bf #1}}
\newcommand{\expn}[1]{{\normalfont\it #1}}
\begin{figure*}
\texttt{\small%
\begin{description}
\setlength{\parskip}{0em}
\setlength{\itemsep}{0em}
\setlength{\labelsep}{0em}
\item[] package harpoon.IR.QuadSSA;
%\item[] import harpoon.ClassFile.*;
\item[]
\item[] \expn{Quadruple statements}
\item[] abstract class Quad implements HCodeElement
\item[\cls{AGET}](HCodeElement source, Temp dst, Temp objectref, Temp index)
\item[\cls{ALENGTH}](HCodeElement source, Temp dst, Temp objectref)
\item[\cls{ANEW}](HCodeElement source, Temp dst, HClass hclass, Temp dims[])
\item[\cls{ASET}](HCodeElement source, Temp objectref, Temp index, Temp src)
\item[\cls{CALL}](HCodeElement source, HMethod method, Temp objectref, 
                  Temp params[], \\ Temp retval, Temp retex) 
                  \expn{// objectref, retval may be null}
\item[\cls{CJMP}](HCodeElement source, Temp test)
\item[\cls{COMPONENTOF}](HCodeElement source, Temp dst, Temp arrayref,
                         Temp objectref)
\item[\cls{CONST}](HCodeElement source, Temp dst, Object value, HClass type)
\item[\cls{FOOTER}](HCodeElement source)
\item[\cls{GET}](HCodeElement source, Temp dst, HField field, \\ 
	         Temp objectref)  \expn{// objectref may be null}
\item[\cls{HEADER}](HCodeElement source, FOOTER footer)
\item[\cls{INSTANCEOF}](HCodeElement source, Temp dst, Temp src, HClass hclass)
\item[\cls{METHODHEADER}](HCodeElement source, FOOTER footer, Temp params[])
\item[\cls{MOVE}](HCodeElement source, Temp dst, Temp src)
\item[\cls{NEW}](HCodeElement source, Temp dst, HClass hclass)
\item[\cls{NOP}](HCodeElement source)
\item[\cls{OPER}](HCodeElement source, String opcode, Temp dst, 
                  Temp operands[])
\item[\cls{PHI}](HCodeElement source, Temp dst[], int arity)
\item[\cls{PHI}](HCodeElement source, Temp dst[], Temp src[][], int arity)
\item[\cls{RETURN}](HCodeElement source, Temp retval) 
                   \expn{// retval may be null}
\item[\cls{SET}](HCodeElement source, HField field, Temp objectref, \\ 
                 Temp src) \expn{// objectref may be null}
\item[\cls{SWITCH}](HCodeElement source, Temp index, int keys[])
\item[\cls{THROW}](HCodeElement source, Temp throwable)
\item[]
\item[] \expn{String constants for} opcode \expn{field of} \cls{OPER}
\item[] \{  "acmpeq", "d2f", "d2i", "d2l", "dadd", "dcmpg",
"dcmpl", "ddiv", "dmul", "dneg", "drem", "dsub", "f2d", "f2i", "f2l",
"fadd", "fcmpg", "fcmpl", "fdiv", "fmul", "fneg", "frem", "fsub",
"i2b", "i2c", "i2d", "i2f", "i2l", "i2s", "iadd", "iand", "icmpeq",
"icmpge", "icmpgt", "idiv", "imul", "ineg", "ior", "irem", "ishl",
"ishr", "isub", "iushr", "ixor", "l2d", "l2f", "l2i", "ladd", "land",
"lcmpeq", "lcmpge", "lcmpgt", "ldiv", "lmul", "lneg", "lor", "lrem",
"lshl", "lshr", "lsub", "lushr", "lxor"
\item[] \};
\end{description}
}

\caption{Classes comprising the QuadSSA intermediate representation.
Only the constructors are shown; the object field variables correspond
exactly to the names of the constructor arguments.}
\label{fig:classes}
\end{figure*}

\bibliography{harpoon}

\onecolumn\appendix\section{Quick reference}
%\begin{figure*}[!p]
\label{sec:quads}
\begin{tabular}{p{1.25in}p{4.5in}} 
\small\bf Class Name & \small\bf Description \\ \hline
\tt HEADER & Start node for control-flow graph.  Performs no operation.\\
\tt METHODHEADER & Subclass of \texttt{HEADER} with additional 
                   method-argument information.\\
\tt FOOTER & End node for control-flow graph.  Performs no operation.\\ \hline
\tt AGET & Fetch from an indexed array element.\\
\tt ALENGTH & Access the length of an array. \\
\tt ANEW & Create a new array object (uninitialized). \\
\tt ASET & Assign a value to an indexed array element. \\
\tt CALL & Invoke an object method. \\
\tt CJMP & Conditional jump based on a boolean value. \\
\tt COMPONENTOF & Determine whether an object reference is an instance
          of the component type of an array reference; result is a boolean
          value. \\
\tt CONST & Assign numeric or string constants to 
            compiler temporary variables. \\
\tt GET & Fetch the value of an object field. \\
\tt INSTANCEOF & Determine whether an object reference is an instance of a
                 given class; result is a boolean value. \\
%MONITOR 
\tt MOVE & Assign one compiler temporary to another. \\
\tt NEW  & Create a new uninitialized class object. \\
\tt NOP  & Do nothing. \\
\tt OPER & Perform a $n$-ary operation on set of compiler temporaries.\\
\tt PHI  & \phifunction{} representation.\\
\tt RETURN & Return a value for this method invocation.\\
\tt SET  & Assign a value to an object field. \\
\tt SWITCH & Jump to one of multiple targets, depending on a key value. \\
\tt THROW & Return an exception for this method invocation. \\
\hline
\end{tabular}
%\caption{Class names of QuadSSA statement types.
%\small (Public classes in \texttt{harpoon.IR.QuadSSA})}
%\end{figure*}

\end{document}
