% -*- latex -*- This is a LaTeX document.
% $Id: thesis.tex,v 1.2 1999-03-20 06:50:53 cananian Exp $
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[notitlepage,twoside]{article}
%\documentclass[12pt,notitlepage,twocolumn,twoside]{article}

% the name of the project
\newcommand{\magic}{\textsc{Magic}}

% A breaking non-space for URLs
\newcommand{\bsp}{\hspace{0pt}}

% phi and sigma functions
\newcommand{\phifunction}{$\phi$-function}
\newcommand{\sigfunction}{$\sigma$-function}

% Cytron's path symbol
\newcommand{\path}{\ensuremath{\rightarrow}}
\newcommand{\pathplus}{\ensuremath{\stackrel{+}{\path}}}

% double-space
\renewcommand{\baselinestretch}{1.2}

\title{\magic: An Optimizing Java Compiler using SSI Form}
\author{C. Scott Ananian}
\date{\today \\ $ $Revision: 1.2 $ $}

\begin{document}
\pagestyle{myheadings}\markboth{$ $Revision: 1.2 $ $}{$ $Revision: 1.2 $ $}
\bibliographystyle{plain}

\maketitle

\section{Introduction}
The choice of compiler internal representation is fundamental.
Advances in IRs translate to advances in compilers.
This paper represents another advance in IRs.

\section{Intermediate Representations}
\section{Static Single Assignment Form}
Static Single Assignment (SSA) form, introduced by Cytron in
\cite{cytron89:ssa}%
\footnote{There was some reference to an earlier use? Where's that reference?}
\ldots

\textbf{Define SSA form here.}

\subsection{Minimal SSA form}
Cytron defines a minimal SSA form in \cite{cytron91:ssa}.  His minimal
form is defined to use the smallest number of \phifunction{s} such
that the following three conditions hold:
\begin{enumerate}
\item If two nonnull paths $X \pathplus Z$ and $Y \pathplus Z$
converge at a node $Z$, and nodes $X$ and $Y$ contain assignments to
[a variable] $V$ (in the original program), then a trivial
\phifunction{} $V \leftarrow \phi(V, \ldots, V)$ has been inserted at
$Z$ (in the new program). \label{criteria1}
\item Each mention of $V$ in the original program or in an inserted
\phifunction{} has been replaced by a mention of a new variable $V_i$,
leaving the new program in SSA form.
\item Along any control flow path, consider any use of a variable $V$
(in the original program) and the corresponding use of $V_i$ (in the
new program).  Then $V$ and $V_i$ have the same value.
\end{enumerate}
Of these criteria, \ref{criteria1} is the most important.  A variation
on minimal SSA form, called \textit{pruned} form \cite{ferrante91:pruned},
avoids placing \phifunction{s} which define variables which are never used.

\section{Static Single Information Form}

\textbf{Define SSI form here.}

\subsection{Minimal SSI form}
\textit{Minimal} and \textit{pruned} SSI forms can be defined which
parallel their SSA counterparts.  \textit{Minimal} SSI form would have the
smallest number of $\phi$- and \sigfunction{s} such that:
\begin{enumerate}
\item If two nonnull paths $X \pathplus Z$ and $Y \pathplus Z$
exist having only the node $Z$ where they converge in common,
and nodes $X$ and $Y$ contain either assignments to a variable $V$ in the
original program or a $\phi$- or \sigfunction{} for $V$ in the new program,
then a \phifunction{} for $V$ has been inserted at $Z$ in the new program.
\item If two nonnull paths $Z \pathplus X$ and $Z \pathplus Y$
exist having only the node $Z$ where they diverge in common,
and nodes $X$ and $Y$ contain either uses of a variable $V$ in the
original program or a $\phi$- or \sigfunction{} for $V$ in the new program,
then a \sigfunction{} for $V$ has been inserted at $Z$ in the new program.
\end{enumerate}

A \textit{pruned} SSI form would be the minimal form without any unused
definitions, that is, $\phi$ or \sigfunction{s} after which there are no
subsequent uses of any of the variables defined on the left-hand side,
except in other $\phi$ or \sigfunction{s}.

\section{Results}
\section{Conclusions}

\bibliography{harpoon}
\appendix
\end{document}
