\documentclass{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{dsfont}
\usepackage{epsfig}
\usepackage{bbold}

\title{}
\author{Instructions}
\date{July 30, 2003}

\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\textwidth}{6.6in}
\setlength{\textheight}{9in}
\setlength{\parindent}{0.5in}



\begin{document}
\maketitle
\begin{flushleft}

\thispagestyle{empty}



\vspace{0.1in}
The goal of this experiment is to evaluate the applicability of our system 
as a debugging tool.  There are two populations taking part in this
experiment.  Both of them are given a series of short C programs (around 500 
lines each) and are asked to find and fix the bugs they contain. The first 
population is given the buggy programs alone, while the second population is 
given the buggy programs and our tool. 

\vspace{0.2in}
You are a member of the ................ population. 

\vspace{0.2in}

More precisely, you will receive a program that implements a simplified version
of the Linux ext2 file system.  A brief description of the ext2 file system can
be found in an attached document. 

\vspace{0.1in}
You will receive three variants of this program.  These variants are extremely
similar, but there might be small differences between them.  You are going
to see one variant at a time, and {\em you are not allowed to go back to work 
on a previous variant}.

\vspace{0.1in}
All three variants are buggy and you are supposed to find and fix the bugs.
Note that the bugs are not given in order of difficulty.  So if after a 
certain amount of time you don't find the first bug, please continue with
the second program.

\vspace{0.1in}
For each buggy program, we ask you to provide the following pieces of 
information:

\begin{enumerate}
\item{The time spent to localize the bug at a procedure level (that is, how 
      much time you spent until you found the procedure that contains the 
      bug);}

\item{The time spent to understand what the bug does (what effect it has on the
      program);}

\item{The time spent to fix the bug.}
\end{enumerate}


\end{flushleft}
\end{document}