%%%% other text which doesn't go anywhere yet

\subsecput{os}{Operating systems}

Perhaps the most promising single application for boundless
transactions is providing atomicity within an operating system.
Locking is difficult to reason about For example, we counted at the
comments in the linux filesystem layer, and found that about 15\% of
the comments relate to locking protocols.  Those comments often
describe global invariants of the program which are difficult to
verify.  And many kernel bugs involve races and deadlocks.

[Dealing with nested transactions: modularity, semantics.]

\subsecput{htm}{HTM versus STM}

\subsecput{performance}{Performance improvements}

\note{This text may very well belong in some other section.}  Herlihy
and Moss' HTM scheme is not non-blocking.  In their proposal, a
transactional load by a processor ends with an XABORT and an XCOMMIT
tag on the read cache line.  Later, while that processor's transaction
is still outstanding, a transactional store by a different processor
to that same cache line will result in a T\_RFO cycle on the bus, to
which the first processor's cache will result BUSY.  The second
processor commits suicide (aborts itself) in response.

[A processor which does transactional reads holds an irrevocable lock
on those cache lines until it commits or aborts itself, and any
other processor which attempts a write on those lines will block,
waiting.]

This means that a processor which transactionally reads many locations
and then loops forever without committing may prevent all other
processors from ever making progress on their transactions.  This is a
simple protocol error: the intervening processor should be able to
abort the first processor instead of blocking for its completion.
Loosely, processors must be able to kill others; if a processor can
only commit suicide, than the system may block.  When transactions are
exposed to user code, Herlihy's system is vulnerable to this simple
denial-of-service attack.  BTM does not suffer from this
problem. [although starvation may still occur.]


